Research Experience

### 2010-04-20 doc/thesis/thesis.tex
some work on the lit review portion

\begin{document}
\section{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity deep in the brain. 
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis is straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite the fact that FMRI is so widely used,
the choice of analysis tools is still relatively limited. Every widely
available analysis tool is based on parametric modeling, which requires
prior knowledge of the noise distribution. Indeed, the linear methods
usually applied are known not to be robust to non-gaussian, non-white
noise, which is why so many pre-processing steps are necessary.
These limitations cast a long shadow over any experiment that claims to be
able to reject the null hypothesis, no matter the P-value. While obviously
there is a correlation between the BOLD signal and actual activity, its
possible the current parametric tests are over-estimating the actual amount
of activity. Its also possible that the current methods underestimate
actual activity. In this paper we propose a robust model-based approach to 
activation detection that makes no assumptions about the underlying noise
model.

fact, typical activation detection does little to account for variations
in the shape of the signal, but instead is a direct T-test between the measured signal
and the expected shape of the signal. Because of these limitations,
multiple comparison tests are imperative (ref fish), and active regions
must be very large to be detected with any confidence. This paper
will introduce a more robust method of analyzing FMRI signals for 
activation, by means of estimating the BOLD parameters.

\subsection{FMRI}
Before going further, it is necessary to discuss Functional MRI,
and what it measures. All MRI works by exciting nuclear spin away
from a steady state controlled by a large electromagnet. Different 
nuclei respond to RF impulses based on their natural frequency
of oscillation. Therefore contrast in MRI is based on the aggregate
nuclear composition of a given region. Thus, regions with more oxygen
will respond differently from regions with proportionately less 
oxygen. This exact difference is what causes the changes over time
in the MR signal in functional MRI. Unfortunately, it takes a significant
amount of time for nuclear spin to return to equilibrium. Thus instead
of waiting, a type of imaging called Echo Planar Imaging (EPI) is used,
wherein a single excitation is used, and then as quickly as possible
every voxel (3 Dimensional Pixel) in a particular region is read. 
Because "reading" a particular region requires the switching on of a
gradient magnetic field, it takes a decent amount of time to perform
measurements. Problematically, the signal from the original excitation
drops exponentially, thus the signal to noise ratio can be very low
even at modest resolutions. Eventually the signal becomes unreadable,
and a new excitation must be applied. As more measurements are made, and thus
as the spatial resolution increases, the rate at which entire volumes
can be acquired drops. Typically volumes can be acquired every two to
three seconds, at resolutions of about 64 by 64 by 32, but their signal
to noise ratio is often poor. Other imaging modalities such as EEG and
MEG are capable of much faster measurements, but are unable to isolate
measurement locations so precisely as FMRI. As with all biological systems,
noise is much higher than in human designed systems, and yet these 
biological systems are capable of performing amazing feats.

@@

>>>>

\begin{document}
\section{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity for the majority of the brain.
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite its prevalence, there have been relatively few developments
in the actual analysis of FMRI images. A steady stream of studies have adapted
the BOLD model from the initial Baloon model first proposed by \cite{Buxton1998}
all the way to full fully autonomous system of equations \cite{Riera2004}. And while
there have been numerous forks in the model, enough in fact to make an entire paper
studying the differences,\cite{Deneux2006}, it is widely known that all these
models have quantitatively less bias error than General Linear Model which is
typically employed today. Then again, depending on the model there may between
seven \cite{Riera2004} and fifteen \cite{Zheng2002} parameters per voxel to
be optimized. Clearly there is a significant risk of error due to variance
with so many degrees of freedom, not to mention a significantly increased
computation cost. In this thesis I demonstrate the use of a 
particle filter as a means of addressing these problems.

\subsection{FMRI}
Magnetic Resonance Imaging, MRI is a method of building 3D images
non-invasively, based on the difference between nuclear spin
relaxation times in various molecules. Initially the entity being
imaged is brought into a large magnetic field which aligns the spins
of molecules in the same direction; radio frequency (RF) signals may
then be used to excite nuclear spin away from the steady alignment. 
As the nuclei precess back to their original orientation, they resonate
at the same RF frequency of their original excitation. Conveniently, the
excitation of nuclear spins return their original state at different
rates, called the T1 relaxation time, depending on the properties of 
the material excited. Additionally, the
coherence of the spins also decay differently (and quite a bit faster
than T1) based on the properties of the region that has been excited.
This gives two primary methods of contrasting substances,
which is the basis of T1 and T2 weighted images. Additionally, there
dephasing occurs at two different rates, the T2 relaxation time,
which is impossible to recover from, and T2$^*$ relaxation, which is
much faster, but possible to recover from with an inversion pulse.
Oftentimes T1 relaxation times can be on the order of seconds if 
a significant excitation pulse is applied. 
In order to rapidly acquire entire brain images, as is done in Functional 
MRI, a single large excitation pulse is applied to the entire brain,
and the entire volume is acquired in a single T1 relaxation period. 
Because the entire k-space (spatial-frequency) volume is acquired 
from a single excitation, the signal to noise ration is very low
in this type  of imaging (Echo Planar Imaging).

Increasing the spatial resolution of EPI imaging necessarily 
requires more time or faster magnetic field switching. Increasing
magnet switching rates though is difficult, because it can result in
more artifacts, or even lower signal to noise ratios. The result is
that at best FMRI is capable of 1.2 second temporal resolution. 
Additionally, the means that each voxel of the image will contain 
the sum of a large amount neurons, capillaries and veins. Thus, the
FMRI signal, which is sensitive to the chemical composition of 
materials, is summing up the composition of various types of tissue
in addition to the blood, whose composition is what we actually care about.
In particular, the presence of Deoxyhemoglobin, Hemoglobin whose
oxygen has been used by a metabolic process, has a larger magnetic
response compared to Oxygenated Hemoglobin. Thus, capillaries near
very active cells will typically have a higher Deoxyhemoglobin content,
and regions with lower activity will have a lower Deoxyhamoglobin content.
Unfortunately, as mentioned previously, blood is only a small part of
each voxel, which means that a single EPI image doesn't tell much about
Deoxyhemoglobin content. However assuming blood is the only thing changing
in the short term, percent difference from a baseline signal \emph{will}
tell us something. FMRI analysis is thus necessarily performed on the percent change
from the baseline. Luckily the assumption that tissue content does not change in
the short run is actually pretty good, although other facts can pollute
the baseline signal, as we will discuss later.

@@

%trash
%stream of new developments in modeling the BOLD signal, nothing has have been
%a plethora of new studies pushing advancements 
%the choice of analysis tools is still relatively limited. Every widely
%available analysis tool is based on parametric modeling, which requires
%prior knowledge of the noise distribution. Indeed, the linear methods
%usually applied are known not to be robust to non-gaussian, non-white
%noise, which is why so many pre-processing steps are necessary.
%These limitations cast a long shadow over any experiment that claims to be
%able to reject the null hypothesis, no matter the P-value. While obviously
%there is a correlation between the BOLD signal and actual activity, its
%possible the current parametric tests are over-estimating the actual amount
%of activity. Its also possible that the current methods underestimate
%actual activity. In this paper we propose a robust model-based approach to 
%activation detection that makes no assumptions about the underlying noise
%model.
